Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-28 text/back_end_KS.qmd
Dallas results smoothing

@@

From 1980 to 2020, these demographic shifts have transformed the Chicago metropolitan area's Geomarket landscape from largely a binary structure into a more tiered dynamic shaped by racial/ethnic and socioeconomic characteristics. The original Geomarket geographical boundaries delineated significant racial and income segregation in the 1980s. This included Black, Hispanic, and Asian residents primarily in the City of Chicago and White residents in the suburban ring of the metropolitan area. Income, poverty, and educational attainment also created a static socioeconomic divide in 1980, with southern Geormarkets (City of Chicago and South and Southwest Suburbs) characterized by lower affluence and educational attainment compared to the higher socioeconomic standing of the northwestern suburbs. However, declining White populations, Hispanic and Asian population growth, and Black outmigration, as well as some narrowing of education and poverty gaps, have contributed to a more stratified Geomarket dynamic in the metropolitan area despite significant disparities remaining. The City of Chicago Geomarket remains the critical geographical hub across all Geomarkets for racial/ethnic diversity; however, other Geomarkets include substantive numbers of residents across racial/ethnic groups (such as South and Southwest Geomarket for Black residents, Chain of Lakes for Hispanic residents, and Northwest Suburbs for Asian residents) while the the Evanston and North Shore Geomarkets remain predominantly White. While some stark socioeconomic gaps within Geomarkets have narrowed since 1980, significant geographical income segregation remains across Geomarkets despite a more clearly defined middle-to-high socioeconomic Geomarket group (Chain of Lakes, Northwest Suburbs, Western Suburbs).

@@

@@

The six Geomarkets within the Dallas-Fort Worth metropolitan area are illustrated in @fig-dallas-maps by total population across the three census periods. The Dallas-Fort Worth region represents a unique metroplex case formed by the convergence of two separate metropolitan areas, which creates a unique urban spatial structure that differs from the classic model originally proposed by @RN4982. As depicted in @fig-dallas-maps, two distinct inner-city core Geomarkets (City of Dallas and City of Fort Worth) are seperated by a centrally located suburban Geomarket encompassing Irving, Arlington, & Grand Prairie. These central Geomarkets are surrounded by the more conventional outer ring of suburban Geomarkets (Dallas County, Collin and Rockwall Counties, and West of Dallas/Ft. Worth Metroplex).

>>>>

### Shifting Demographics, Persistent Segregation in Geomarket Landscape

@@

From 1980 to 2020, these demographic shifts have transformed the Chicago metropolitan area's Geomarket landscape from largely a binary structure into a more tiered dynamic shaped by racial/ethnic and socioeconomic characteristics. The original Geomarket boundaries delineated significant racial and income segregation in the 1980s. This included Black, Hispanic, and Asian residents primarily in the City of Chicago Geomarket and White residents in Geomarkets forming the suburban ring of the metropolitan area. Income, poverty, and educational attainment also created a static socioeconomic divide in 1980, with southern Geormarkets (City of Chicago and South and Southwest Suburbs) characterized by lower affluence and educational attainment compared to the higher socioeconomic standing of the northwestern suburbs. However, declining White populations, Hispanic and Asian population growth, and Black outmigration, as well as some narrowing of education and poverty gaps, have contributed to a more stratified Geomarket dynamic in the metropolitan area despite significant disparities remaining. The City of Chicago Geomarket remains the critical geographical hub across all Geomarkets for racial/ethnic diversity; however, other Geomarkets include substantive numbers of residents across racial/ethnic groups (such as South and Southwest Geomarket for Black residents, Chain of Lakes for Hispanic residents, and Northwest Suburbs for Asian residents) while the the Evanston and North Shore Geomarkets remain predominantly White. While some stark socioeconomic gaps within Geomarkets have narrowed since 1980, significant geographical income segregation remains across Geomarkets despite a more clearly defined middle-to-high socioeconomic Geomarket group (Chain of Lakes, Northwest Suburbs, Western Suburbs).

### Composition of Students Lists by Geomarket Contributions

@@

The enduring segregated structure of the Chicago metropolitan area from 1980 to 2020 analyzed above plays a significant role in how Geomarkets shape the racial/ethnic and socioeconomic composition of prospect pools. Reflecting the larger metropolitan area dynamics, more central Geomarkets such as the City of Chicago and South and Southwest Suburbs consistently contribute disproportionately larger shares of Black, Hispanic, and first-generation college prospects, especially at lower SAT score thresholds. On the other hand, Geomarkets in the suburban ring of the metropolitan area like the Chain of Lakes, North Shore, and Evanston and Skokie contribute larger proportions of White and Asian prospects, particularly as SAT scores increase. However, the City of Chicago stands out in playing crucial role in fostering college access for first-generation Asian college students, contributing such prospects to the overall pool at significantly higher, disproportionate rates than any other Geomarket.

@@

### Suburban Geomarket Growth Eases Racial Disparities, Concentrated Wealth Persists

The six Geomarkets within the Dallas-Fort Worth metropolitan area are illustrated in @fig-dallas-maps by total population across the three census periods. The Dallas-Fort Worth region represents a unique "metroplex" case formed by the convergence of two separate metropolitan areas, which creates a unique urban spatial structure that differs from the classic urban core and suburban ring model showcased by Chicago. As depicted in @fig-dallas-maps, two distinct inner-core Geomarkets (City of Dallas and City of Fort Worth) are seperated by a centrally located suburban Geomarket encompassing Irving, Arlington, & Grand Prairie. These central Geomarkets are surrounded by the more conventional outer ring of suburban Geomarkets (Dallas County, Collin and Rockwall Counties, and West of Dallas/Ft. Worth Metroplex).
